Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-16 analysis/paper/paper.Rmd

The Deadwater Slough is located in a reach of the Salmon River that is believed to be a historically important overwinter rearing area for DSR emigrants. Moreover, this reach is part of the migratory pathway for juvenile DSR and NRR emigrants from all eight extant populations. The slough lacks hydrological and structural features (i.e., a homogenous channel with fine substrate and little cover) that can provide essential refuge from predation. As a result, predation on juvenile Chinook salmon proximal to the Deadwater Slough has been cited as a concern for the Upper Salmon River MPG, impacting DSR migrants in the fall and NRR emigrants during the spring.

We hypothesize that increased densities of piscivorous predators in the Deadwater Slough area may explain the reduced survival (or apparent survival) observed for juvenile Chinook Salmon [@Ackerman2018] and Sockeye Salmon [@Axel2015]. In this study, we estimated the abundance of a piscivorous fish predator population in the Deadwater Slough and their potential impacts to juvenile salmon emigrants, focusing on DSR and NRR Chinook Salmon. Our objectives for the study were four-fold:

@@

We estimated the population size of Northern Pikeminnow in the Deadwater Slough to be greater than ***xx,xxx*** during the fall emigration period for DSR Chinook Salmon. That estimate translates to a density of ***xxx*** Northern Pikeminnow per 100 m or ***xxx*** per 100 m^2^ which is ***similar/more/less*** than estimates from elsewhere in the Columbia River ***(citation)*** where substantial Northern Pikeminnow predation impacts on salmonids have led to bounty programs aimed at reducing Northern Pikeminno abundance. The population size of Northern Pikeminnow was not directly estimated during the spring NRR Chinook salmon emigration period however, the relative abundance measured at CPUE was comparable to the fall sampling periods ***(update statement later)***. The population of Northern Pikeminnow in Deadwater Slough was estimated to consume between ***xx,xxx*** and ***xx,xxx*** juvenile Chinook Salmon during the ***x*** sampling periods and result in an estimated reduction of returning adults between ***xxx*** and ***x,xxx***. We suggest that the habitat modifications that created the Deadwater Slough reach have resulted in favorable conditions for Northern Pikeminnow, including improved conditions for predation upon juvenile Chinook Salmon ***(add detail here)***. Therefore, predation by Northern Pikeminnow in the Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG.

@@

>>>>

The Deadwater Slough is in a reach of the Salmon River that is believed to be a historically important overwinter rearing area for DSR emigrants. Moreover, this reach is part of the migratory pathway for juvenile DSR and NRR emigrants from all eight extant populations. The slough lacks hydrological and structural features (i.e., a homogenous channel with fine substrate and little cover) that can provide essential refuge from predation. As a result, predation on juvenile Chinook salmon proximal to the Deadwater Slough has been cited as a concern for the Upper Salmon River MPG, impacting DSR migrants in the fall and NRR emigrants during the spring.

We hypothesize that increased densities of piscivorous predators in the Deadwater Slough may explain the reduced survival (or apparent survival) observed for juvenile Chinook Salmon [@Ackerman2018] and Sockeye Salmon [@Axel2015]. In this study, we estimated the abundance of a piscivorous fish predator population in the Deadwater Slough and their potential impacts to juvenile salmon emigrants, focusing on DSR and NRR Chinook Salmon. Our objectives for the study were four-fold:

@@

We estimated the population size of Northern Pikeminnow in the Deadwater Slough to be greater than ***xx,xxx*** during the fall emigration period for DSR Chinook Salmon. That estimate translates to a density of ***xxx*** Northern Pikeminnow per 100 m or ***xxx*** per 100 m^2^ which is ***similar/more/less*** than estimates from elsewhere in the Columbia River ***(citation)*** where substantial Northern Pikeminnow predation impacts on salmonids have led to bounty programs aimed at reducing Northern Pikeminno abundance. The population size of Northern Pikeminnow was not directly estimated during the spring NRR Chinook salmon emigration period however, the relative abundance measured at CPUE was comparable to the fall sampling periods ***(update statement later)***. The population of Northern Pikeminnow in Deadwater Slough was estimated to consume between ***xx,xxx*** and ***xx,xxx*** juvenile Chinook Salmon during the ***x*** sampling periods and result in an estimated reduction of returning adults between ***xxx*** and ***x,xxx***. We suggest that the habitat modifications that created the Deadwater Slough have resulted in favorable conditions for Northern Pikeminnow, including improved conditions for predation upon juvenile Chinook Salmon ***(add detail here)***. Therefore, predation by Northern Pikeminnow in the Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG.

@@

<!--Please feel free to add others that I missed or improve*** -->

The authors extent much appreciation to the many volunteers who assisted with field efforts including collaborators from Bureau of Reclamation, Idaho Department of Fish and Game, and Lemhi Regional Land Trust, among others. This manuscript benefited from reviews and contributions from colleagues at the Idaho Governor's Office of Species Conservation, Rio Applied Science and Engineering, and from Sean Gibbs and Ben Briscoe at Mount Hood Environmental. Funding for this study was provided by the Bureau of Reclamation, Pacific Northwest Regional Office (contract No. 140R1021F0018). Special thanks to Caitlin Alcott and Inter-Fluve for their administrative support and guidance.
